Prediction of PIK3CA mutation with gene expression

### 2020-05-07 plosOne/plosOne.Rmd
edit 5/7

Targeted thrapy becomes standard treatment in many cancer patients. To treat with the targeted therapy, the The patients should be confirmed the indicated genomic alterations in their cancer. Many direct test for the genomic alteration has been developed and prooved for their clical utility to discriminate which patients will be response the targeted therapy.

The machine learning approach has been actively researched recently to detect the genomic alterations. The machine learning can build a prediction model from a large number of predictors such as radiolomic features @dercleIdentificationNonSmall2020, pathology image or gene expression data @wayMachineLearningDetects2018. The machine learning approach has limitations to apply to clinical practice. However the machine learning prediction can be a second best when the direct test fails.

@@

Prediction RAS pathway activation with gene expression data was done in previous study. @way_2018_machine_cellreports They trained pancancer The Cancer Genome Atlas (TCGA) data with a supervised elastic net penalized logistic regression classifier with stochastic gradient descent. The performance of their model was 84% with an area under the receiver operating characteristic (AUROC) curve and 63% with an area under the precision recall (AUPR) curve. We applied their modeling methods in prediction PIK3CA mutation.

@@

10845 cases were available both PIK3CA mutation and mRNA expression data. 5128 out of 20502 genes were included in the modeling process after filtering with median absolute deviation as described at modelinf process method. 33 cancer type dummy variables were included in predictor variables.

The target variable was PIK3CA mutation status. The status of  PIK3CA was considered as positive when the case has following PIK3CA variants which is the target variables of the therascreen PIK3CA RGQ PCR Kit; C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R, H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved by U.S. Food and Drug adminidtration for PIK3CA inhibitor. We splited the three quarters of dataset for the trainset and one quarter for testset.

@@

## Dataset summary 
Prevalcence of PIK3CA was 0.11 in all cases. The PIK3CA prevalance of each cancer type was vary. The median prevalence of PIK3CA of each cancer types  was 0.03 (range 0-0.33) (Figure 1).

@@

In 10-fold cross-validation, the model with lambda = 0.01 and alpha = 1.0 (Ridge regression) were best performance in terms of AUROC. The final model was trained with the selected hyperparameters with all trainset.

@@

Our model showed good performance to predict PIK3CA mutation of various cancer types.

@@

@@

OUr model includes cancer type predictor and they are stronger than gene expression data. The varing prevalence of PIC3CA mutation across cancer type might reason of the strong cancer type predictor. If the cancer type was wrong or can not be determined, our model performance can be poor.

@@

- Figure 1. Summary of modeling results

@@

- Figure 2. Coefficients of model

(A) Top 30 high coefficients of mRNA. (B) Coefficients of cancer types. The abbreviations are explained in S1 appendix.

>>>>

Targeted thrapy becomes standard treatment in many cancer patients. Many targeted therapy requires test for a specific cancer genomic alteration, to treat the patients. Many direct test for the genomic alteration has been developed and prooved for their clical utility to discriminate which patients will be response the targeted therapy.

Machine learning approach has been actively researched to detect the genomic alterations. Machine learning can build a prediction model from a large number of predictors such as radiolomic features @dercleIdentificationNonSmall2020, pathology image or gene expression data @wayMachineLearningDetects2018. Because most direct genomic test are more specific and sensitive than predictive models, machine learning approach might has limited role in clinical practice. However the machine learning prediction can be a second best when the direct test fails.

Prediction RAS pathway activation with gene expression data was done in previous study. @way_2018_machine_cellreports They trained pancancer The Cancer Genome Atlas (TCGA) data with a supervised elastic net penalized logistic regression classifier with stochastic gradient descent. The performance of their model was 84% with an area under the receiver operating characteristic (AUROC) curve and 63% with an area under the precision recall (AUPR) curve. The authors suggested their approach can be applied to other genomic alterations.

@@

We apply a supervised elastic net penalized logistic regression model in prediction PIK3CA mutation. The purpose of this study is to know this prediction model approach can be applied not only RAS pathway activation but also PIK3CA mutation across many cancer types.

@@

10845 cases were available both PIK3CA mutation and mRNA expression data. 5128 out of 20502 genes were included in the modeling process after filtering with median absolute deviation as described at modeling process method. 33 cancer type dummy variables were included in predictor variables.

The target variable was PIK3CA mutation status. The status of  PIK3CA was considered as positive when the case has following PIK3CA variants which is the target variables of the therascreen PIK3CA RGQ PCR Kit; C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R, H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved as a companion diagnosis to treat with PIK3CA inhibitor by U.S. Food and Drug adminidtration. We splited the three quarters of dataset for the trainset and one quarter for testset.

@@

## prevalance rate rate of PIK3CA mutation   
prevalance rate rate of PIK3CA was 0.11 in all cases. The PIK3CA prevalance rate of each cancer type was vary. The median prevalence of PIK3CA of each cancer types  was 0.03 (range 0-0.33) (Figure 1).

@@

In 10-fold cross-validation, the model with $\lambda$ = 0.01 and $\alpha$ = 1.0 (Ridge regression) showed best performance in terms of AUROC. The final model was trained with the selected hyperparameters with all trainset.

@@

@@

Our model showed good performance to predict PIK3CA mutation of various cancer types. This result shows that the supervised elastic net penalized logistic regression model can be applied not only RAS activation pathway but also PIK3CA mutation. Both RAS activation pathway and PIK3CA mutation are important and common cancer genomic alterations. They have significant effect on gene expression in cacer cells. It might be challenging prediction of genomic alterations which are infrequent or have weak effect on gene expression.

@@

Our model includes cancer type predictor and they are stronger than gene expression data. The varing prevalence of PIC3CA mutation across cancer type might reason of the strong cancer type predictor. If the cancer type was wrong or can not be determined, our model performance can be poor.

@@

- Figure 1. prevalance rate rate of PIK3CA across cancer types
The abbreviations of cancer types are explained in S1 appendix.

- Figure 2. Summary of modeling results

@@

- Figure 3. Coefficients of model

(A) Top 30 high coefficients of mRNA. (B) Coefficients of cancer types. The abbreviations of cancer types are explained in S1 appendix.

### 2020-05-07 plosOne/plosOne.tex
edit 5/7

The machine learning approach has been actively researched recently to
detect the genomic alterations. Breast cancer with PIK3CA mutation has
been approved to use PIK3CA inhibitor in hormone receptor positive HER2
negative subtype. {[}1{]}

\begin{itemize}
\tightlist
\item
  PIK3CA summary
\end{itemize}

PIK3CA is encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase which
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to make phosphatidylinositol 3,4,5-triphosphate
(PIP\_\{3\}). Phosphatase and tensin homolog (PTEN) changes PIP\_\{2\}
to PIP\_\{3\} in contrast PI3K. PIP\_\{3\} is a second mesenger to
activate protein kinase B (AKT) which is a serine/threonine-specific
protein kinase. AKT inhibits apoptosis and promote cell proliferation.

The PIK3CA mutation is second most driver mutation after TP53. the
PIK3CA mutation is most frequantly founded in endometrial carcinoma
(45\%), and followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma and endocervical adenocarcinoma (20\%) and colon
adenocarcinoma (16\%).

\begin{itemize}
\tightlist
\item
  Predicting mutation by gene expression data

  \begin{itemize}
  \tightlist
  \item
    What for?\\
  \item
    Feasible?\\
  \item
    Cost?
  \end{itemize}
\item
  Varing frquency of PIK3CA mutation across cancer types

  \begin{itemize}
  \tightlist
  \item
    High in endometrial breast\\
  \item
    not common other cancer type
  \end{itemize}
\end{itemize}

Universial prediction using gene expression data across cancer type for
certain kind of mutation. It's clinically useless now, but we want to
explore the possibilities of the PIK3CA mutation prediction.

\begin{itemize}
\item
  RNAseq can be widely used. The mutation status directly prediction
\item
  Previous study {[}2{]}
\item
  Prediction of aberant activation of a certain pathway vs a specific
  mutation
\item
  The proposed solution
\end{itemize}

Prediction RAS pathway activation with gene expression data was done in
previous study. {[}2{]} They trained pancancer The Cancer Genome Atlas
(TCGA) data with a supervised elastic net penalized logistic regression
classifier with stochastic gradient descent. The performance of their
model was 84\% with an area under the receiver operating characteristic
(AUROC) curve and 63\% with an area under the precision recall (AUPR)
curve. We applied their modeling methods in prediction PIK3CA mutation.

@@

We used TCGA pancancer dataset. PIK3CA mutation data was get using cgdsr
rpackage.{[}3{]} Gene expression data was get from GDAC firehose using
RTCGAToolbox R package. {[}4{]}

@@

To narrow down potential predictors, Genes with a large the median
absolute deviation (more than third-quartiles) were selected.
Yeo-Johnson transformation was done to correct skewness. Centering and
scaling were done. All preprocessing was done using recipe r package.
{[}5{]} Penalized logistic regression was applied to prediction
modeling. 10-fold cross-validation with targe variable stratification
was done over the hyperparameter grid: \(\lambda\) \{\(10^{-5}\),
\(10^{-4}\),\(10^{-3}\),\(10^{-2}\),\(10^{-1}\), \(10^{0}\)\},
\(\alpha\) \{0.0, 0.25, 0.5, 0.75\}. Lambda is penalty scaling parameter
and alpha is mixing parameter of penalty function
(\((1-\alpha)/2 \lVert\beta\rVert_2^2+\alpha\lVert \beta \rVert_1\)).
{[}6{]} Model performance was evealuated with the area under receiver
operating characteristic (ROC) and area under the ROC (AUROC) and
precision recall (PR) and area under curve (AUPR). The modelin process
was done with tidymodels rpackage. {[}7{]}

@@

In this study, the prevalence of PIK3CA mutation was 11\%. The low
prevalence of PIK3CA results in imbalaced dataset which makes the
prediction difficult. The AUROC is optimistic in terms of performance.
The AUPR is more informative than AUROC on imbalaced datasets. {[}8{]}

@@

\begin{itemize}
\tightlist
\item
  Figure 1. Summary of modeling results
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Figure 2. Coefficients of model
\end{itemize}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Top 30 high coefficients of mRNA. (B) Coefficients of cancer types.
  The abbreviations are explained in S1 appendix.
\end{enumerate}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-andre_2019_alpelisib_nengljmeda}{}%
1. André F, Ciruelos E, Rubovszky G, Campone M, Loibl S, Rugo HS, et al.
Alpelisib for PIK3CA-Mutated, Hormone ReceptorPositive Advanced Breast
Cancer. New England Journal of Medicine. Massachusetts Medical Society;
2019;380: 1929--1940.
doi:\href{https://doi.org/10.1056/NEJMoa1813904}{10.1056/NEJMoa1813904}

\leavevmode\hypertarget{ref-way_2018_machine_cellreports}{}%
2. Way GP, Sanchez-Vega F, La K, Armenia J, Chatila WK, Luna A, et al.
Machine Learning Detects Pan-cancer Ras Pathway Activation in The Cancer
Genome Atlas. Cell Reports. Elsevier; 2018;23: 172--180.e3.
doi:\href{https://doi.org/10.1016/j.celrep.2018.03.046}{10.1016/j.celrep.2018.03.046}

\leavevmode\hypertarget{ref-jacobsenCgdsrRbasedAPI2019}{}%
3. Jacobsen A, Luna A. Cgdsr: R-based API for accessing the MSKCC cancer
genomics data server (CGDS). 2019.

\leavevmode\hypertarget{ref-samurRTCGAToolboxNewTool2014}{}%
4. Samur MK. RTCGAToolbox: A new tool for exporting TCGA Firehose data.
PLoS One. 2014;9(9):e106397.

\leavevmode\hypertarget{ref-kuhnRecipesPreprocessingTools2020}{}%
5. Kuhn M, Wickham H. Recipes: Preprocessing tools to create design
matrices. 2020.

\leavevmode\hypertarget{ref-friedmanRegularizationPathsGeneralized2010}{}%
6. Friedman J, Hastie T, Tibshirani R. Regularization Paths for
Generalized Linear Models via Coordinate Descent. J Stat Softw. 2010;33:
1--22.

\leavevmode\hypertarget{ref-kuhnTidymodelsEasilyInstall2020}{}%
7. Kuhn M, Wickham H. Tidymodels: Easily install and load the
'tidymodels' packages. 2020.

\leavevmode\hypertarget{ref-saitoPrecisionRecallPlotMore2015}{}%
8. Saito T, Rehmsmeier M. The Precision-Recall Plot Is More Informative
than the ROC Plot When Evaluating Binary Classifiers on Imbalanced
Datasets. PLOS ONE. Public Library of Science; 10: e0118432.
doi:\href{https://doi.org/10.1371/journal.pone.0118432}{10.1371/journal.pone.0118432}

>>>>

The machine learning approach has been actively researched recently to
detect the genomic alterations. The machine learning can build a
prediction model from a large number of predictors such as radiolomic
features {[}1{]}, pathology image or gene expression data {[}2{]}. The
machine learning approach has limitations to apply to clinical practice.
However the machine learning prediction can be a second best when the
direct test fails.

PIK3CA is encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase which
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to make phosphatidylinositol 3,4,5-triphosphate
(PIP\textsubscript{3}). Phosphatase and tensin homolog (PTEN) changes
PIP\textsubscript{2} to PIP\textsubscript{3} in contrast PI3K.
PIP\textsubscript{3} is a second mesenger to activate protein kinase B
(AKT) which is a serine/threonine-specific protein kinase. AKT inhibits
apoptosis and promote cell proliferation.

Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor positive HER2 negative subtype. {[}3{]}
The PIK3CA mutation is second most driver mutation after TP53. The
PIK3CA mutation is most frequantly founded in endometrial carcinoma
(45\%), and followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma and endocervical adenocarcinoma (20\%) and colon
adenocarcinoma (16\%).

Prediction RAS pathway activation with gene expression data was done in
previous study. {[}4{]} They trained pancancer The Cancer Genome Atlas
(TCGA) data with a supervised elastic net penalized logistic regression
classifier with stochastic gradient descent. The performance of their
model was 84\% with an area under the receiver operating characteristic
(AUROC) curve and 63\% with an area under the precision recall (AUPR)
curve. We applied their modeling methods in prediction PIK3CA mutation.

@@

We used TCGA pancancer dataset. PIK3CA mutation data was get using cgdsr
rpackage.{[}5{]} Gene expression data was get from GDAC firehose using
RTCGAToolbox R package. {[}6{]}

@@

To narrow down potential predictors, Genes with a large the median
absolute deviation (more than third-quartiles) were selected.
Yeo-Johnson transformation was done to correct skewness. Centering and
scaling were done. All preprocessing was done using recipe r package.
{[}7{]} Penalized logistic regression was applied to prediction
modeling. 10-fold cross-validation with targe variable stratification
was done over the hyperparameter grid: \(\lambda\) \{\(10^{-5}\),
\(10^{-4}\),\(10^{-3}\),\(10^{-2}\),\(10^{-1}\), \(10^{0}\)\},
\(\alpha\) \{0.0, 0.25, 0.5, 0.75\}. Lambda is penalty scaling parameter
and alpha is mixing parameter of penalty function
(\((1-\alpha)/2 \lVert\beta\rVert_2^2+\alpha\lVert \beta \rVert_1\)).
{[}8{]} Model performance was evealuated with the area under receiver
operating characteristic (ROC) and area under the ROC (AUROC) and
precision recall (PR) and area under curve (AUPR). The modelin process
was done with tidymodels rpackage. {[}9{]}

@@

In this study, the prevalence of PIK3CA mutation was 11\%. The low
prevalence of PIK3CA results in imbalaced dataset which makes the
prediction difficult. The AUROC is optimistic in terms of performance.
The AUPR is more informative than AUROC on imbalaced datasets. {[}10{]}

@@

\begin{itemize}
\item
  Figure 1. Prevalance rate of PIK3CA across cancer types The
  abbreviations of cancer types are explained in S1 appendix.
\item
  Figure 2. Summary of modeling results
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Figure 3. Coefficients of model
\end{itemize}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Top 30 high coefficients of mRNA. (B) Coefficients of cancer types.
  The abbreviations of cancer types are explained in S1 appendix.
\end{enumerate}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-dercleIdentificationNonSmall2020}{}%
1. Dercle L, Fronheiser M, Lu L, Du S, Hayes W, Leung DK, et al.
Identification of NonSmall Cell Lung Cancer Sensitive to Systemic Cancer
Therapies Using Radiomics. Clin Cancer Res. American Association for
Cancer Research; 2020;26: 2151--2162.
doi:\href{https://doi.org/10.1158/1078-0432.CCR-19-2942}{10.1158/1078-0432.CCR-19-2942}

\leavevmode\hypertarget{ref-wayMachineLearningDetects2018}{}%
2. Way GP, Sanchez-Vega F, La K, Armenia J, Chatila WK, Luna A, et al.
Machine Learning Detects Pan-cancer Ras Pathway Activation in The Cancer
Genome Atlas. Cell Reports. Elsevier; 2018;23: 172--180.e3.
doi:\href{https://doi.org/10.1016/j.celrep.2018.03.046}{10.1016/j.celrep.2018.03.046}

\leavevmode\hypertarget{ref-andre_2019_alpelisib_nengljmeda}{}%
3. André F, Ciruelos E, Rubovszky G, Campone M, Loibl S, Rugo HS, et al.
Alpelisib for PIK3CA-Mutated, Hormone ReceptorPositive Advanced Breast
Cancer. New England Journal of Medicine. Massachusetts Medical Society;
2019;380: 1929--1940.
doi:\href{https://doi.org/10.1056/NEJMoa1813904}{10.1056/NEJMoa1813904}

\leavevmode\hypertarget{ref-way_2018_machine_cellreports}{}%
4. Way GP, Sanchez-Vega F, La K, Armenia J, Chatila WK, Luna A, et al.
Machine Learning Detects Pan-cancer Ras Pathway Activation in The Cancer
Genome Atlas. Cell Reports. Elsevier; 2018;23: 172--180.e3.
doi:\href{https://doi.org/10.1016/j.celrep.2018.03.046}{10.1016/j.celrep.2018.03.046}

\leavevmode\hypertarget{ref-jacobsenCgdsrRbasedAPI2019}{}%
5. Jacobsen A, Luna A. Cgdsr: R-based API for accessing the MSKCC cancer
genomics data server (CGDS). 2019.

\leavevmode\hypertarget{ref-samurRTCGAToolboxNewTool2014}{}%
6. Samur MK. RTCGAToolbox: A new tool for exporting TCGA Firehose data.
PLoS One. 2014;9(9):e106397.

\leavevmode\hypertarget{ref-kuhnRecipesPreprocessingTools2020}{}%
7. Kuhn M, Wickham H. Recipes: Preprocessing tools to create design
matrices. 2020.

\leavevmode\hypertarget{ref-friedmanRegularizationPathsGeneralized2010}{}%
8. Friedman J, Hastie T, Tibshirani R. Regularization Paths for
Generalized Linear Models via Coordinate Descent. J Stat Softw. 2010;33:
1--22.

\leavevmode\hypertarget{ref-kuhnTidymodelsEasilyInstall2020}{}%
9. Kuhn M, Wickham H. Tidymodels: Easily install and load the
'tidymodels' packages. 2020.

\leavevmode\hypertarget{ref-saitoPrecisionRecallPlotMore2015}{}%
10. Saito T, Rehmsmeier M. The Precision-Recall Plot Is More Informative
than the ROC Plot When Evaluating Binary Classifiers on Imbalanced
Datasets. PLOS ONE. Public Library of Science; 10: e0118432.
doi:\href{https://doi.org/10.1371/journal.pone.0118432}{10.1371/journal.pone.0118432}
